---
title: "Paper Overview"
---

## Abstract

Predicting how diseases like malaria spread is a big challenge, especially as climate change shifts global temperature patterns. A key part of this prediction involves understanding how temperature affects different biological traits of mosquitoes and the parasites they carry — for example, how quickly mosquitoes develop (MDR), how fast the malaria parasite matures inside them (PDR), or how often mosquitoes bite humans. These temperature-sensitive traits are used to calculate something called the basic reproduction number, or R_0, which tells us how easily a disease can spread in a population. 

The problem is that reliable data for these traits — especially at different temperatures — is limited and often noisy. Instead of drawing a single best-fit curve through the data, the paper we are following uses a Bayesian approach with a technique called Markov Chain Monte Carlo (MCMC) to model the uncertainty. This method creates a range of possible trait curves, not just one, and gives us a clearer picture of how confident we are in the estimates.

In our project, we recreate part of this modeling framework using publicly available data from the VectorByte platform. We focus on two or three key traits: mosquito development rate (MDR), parasite development rate (PDR), and possibly mosquito bite rate. Using Bayesian methods and JAGS, we fit temperature-performance curves for these traits, visualize their uncertainty, and show how this uncertainty can impact larger models of disease transmission like R_0.

## 1. Data Collection

The authors assembled two types of datasets:

- **Main data**: Laboratory data under constant temperature, focused on Anopheles gambiae and Plasmodium falciparum. These data were only available for a few traits like mosquito development rate (MDR), egg-to-adult survival (pEA), and adult mosquito mortality ($\mu$).
- **Prior data**: A broader set that included similar traits from related species and less controlled experiments. This dataset was used to construct informative priors in the Bayesian models.

### Plasmodium Falciparum
```{r, echo=FALSE, message=FALSE}
library(DT)
library(readxl)
library(readr)
library(here)

# Load both datasets
mosquitoes1 <- read_excel(here("data", "the_one_we_need.xlsx"))
mosquitoes2 <- read_csv(here("data", "the_one_we_need_2.csv"), show_col_types = FALSE)

# Render first table
datatable(
  mosquitoes1,
  options = list(
    pageLength = 5,
    scrollX = TRUE,
    autoWidth = TRUE,
    columnDefs = list(
      list(width = '35px', targets = "_all")
    )
  ),
  class = 'compact stripe hover',
  rownames = FALSE,
  filter = 'top',
  callback = JS(
    "$('table.dataTable').css({'font-size': '13px'});",
    "$('table.dataTable tbody td').css({'padding': '5px'});"
  )
)
```

### Anopheles Gambiae
```{r, echo=FALSE, message=FALSE}
# Render second table
datatable(
  mosquitoes2,
  options = list(
    pageLength = 5,
    scrollX = TRUE,
    autoWidth = TRUE,
    columnDefs = list(
      list(width = '35px', targets = "_all")
    )
  ),
  class = 'compact stripe hover',
  rownames = FALSE,
  filter = 'top',
  callback = JS(
    "$('table.dataTable').css({'font-size': '13px'});",
    "$('table.dataTable tbody td').css({'padding': '5px'});"
  )
)
```

## 2. Modeling Thermal Responses

Each biological trait (component of the basic reproductive number, R₀) was modeled as a **unimodal response to temperature**, typically hump-shaped.

Two types of thermal performance functions were used:

- **Brière Model** (asymmetric, for traits like MDR):
  $$
  f(T) = a \cdot T \cdot (T - T_0) \cdot \sqrt{T_m - T}
  $$

- **Quadratic Model** (symmetric, for other traits):
  $$
  f(T) = -a \cdot (T - T_{opt})^2 + b
  $$

Choice of function was based on the biological properties of each trait.

## 3. Likelihoods and Constraints

- Most traits used a **truncated normal likelihood**, which ensures outputs are biologically feasible 

- For probability traits (like vector competence or survival), the authors used a **binomial likelihood** when raw counts were available.

- All values were constrained within biologically meaningful ranges. For example, traits expressed as probabilities were constrained to [0, 1].

## 4. Bayesian Inference Workflow

Each trait was modeled individually with the following steps:

1. **Define the likelihood** using the selected functional form (Brière or quadratic).
2. **Assign priors**:
   - Start with **uninformative priors** (flat/diffuse).
   - Later use **informative priors** derived from the prior dataset and expert opinion.
3. **Run MCMC** in **JAGS** to sample from the posterior distribution of parameters:
   - Brière: `a`, `T₀`, `Tₘ`, `σ`
4. **Generate fitted curves** using the posterior draws:
   - Median curve (posterior mean)
   - 95% credible interval envelope

